---
title: "Fitting multi-rate Mk models of character evolution"
author: Michael C. Grundler
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting multi-rate models of character evolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=7, fig.align="center")
```

This vignette will walk through a brief example of how we can use the `mk::mk.shift`
function to fit multi-rate models of character evolution to empirical data.

```{r}
library(mk)
# Data on squamate reproductive modes from
# Pyron, R.A. and F.T. Burbrink. 2014. Ecology Letters 17(1): 13-21
data(squamatatree)
data(squamatareprod)
phy = phylo::read.newick(text=squamatatree)
# A named vector of class ‘factor’ with reproductive mode
# classifications for 3962 squamates.
head(squamatareprod)
```

As a first exploration of the data we'll simply plot the distribution
of parsimony-inferred changes.

```{r}
mpr = parsimony::mpr.fitch(phy, squamatareprod)

# The minimum number of changes required to explain the distribution of
# reproductive modes observed at the tips of the tree
mpr$score

# The total number of MPR histories of character state change
mpr$mpr_count
```

For this example we'll work with a single MPR history. The object returned
by `mpr.fitch` includes a `simulate` function to draw MPR histories of character 
state change at random.

```{r}
# This returns a matrix. Each column is a single history and rows represent 
# nodes in the phylogeny. The cell value is the character state assigned to the 
# node whose index is the same as the row index. For 'tree' class phylogenies 
# used in the mk package, node indices are identical to the node indices of an 
# ape 'phylo' object in 'cladewise' node order.
h = mpr$simulate(1)
```

We'll use the results stored in `h` to color branches according to their
inferred reproductive mode. Oviparous (egg-laying) lineages will receive
a blue color and viviparous ("live-bearing") lineages will receive a
red color.

```{r}
edge.color = ifelse(h[,1] == 1L, 4L, 2L)
```

Let's also use `h` to identify all instances of character state change
To do so, we'll traverse the phylogeny and identify where a node's 
inferred character state differs from that of its ancestor

```{r}
state.change = logical(phylo::Nnode(phy))
for (node in phylo::descendants(root(phy), phy))
{
    if (h[node,] != h[phylo::parent(phy, node),])
        state.change[node] = TRUE
}

# Confirm that this equals the number of MPR inferred state changes from before
sum(state.change) == mpr$score
```

Now we're ready for some plotting. Note that the plot function invisibly returns 
the plotting coordinates, which we'll store for making annotations.

```{r}
L = plot(phy, edge.color=edge.color)
title("Maximum parsimony inferred character states")

# Now highlight the nodes that underwent a state change
points(L[[1]][state.change, 1], L[[1]][state.change, 3], pch=21, 
    bg=edge.color[state.change], cex=0.8)
```

We can observe that state changes are definitely clustered in certain
regions of the phylogeny and that changes away from oviparity are much
more common. 

The conspicuous phylogenetic clustering of changes is certainly
suggestive of rate heterogeneity, so let's fit a multi-rate model
of reproductive mode evolution and see how it aligns with the
distribution of inferred character state changes.

This is actually very simple:

```{r}
fit = mk::mk.shift(squamatareprod, phy)
str(fit)
```

This returns a list with three components.

The first component is a vector of lineage-specific rates of character
evolution that have been average over multiple possible rate-shift
configurations. The second component is a set of weights assigned to
different rate-shift configurations. These weights were used to compute
the average rates in the first component. The final component is a function
that can be used to retrieve the lineage-specific rates of character
evolution associated with a particular rate-shift configuration.

Let's first look at the model-averaged rates. Note that each value in
this vector is the rate assigned to the edge leading to the node whose index 
is the same as the vector index.

```{r}
edge.rate = fit$avg.rates
```

Now let's discretize the rates and then map them to a set of colors, where
warmer colors mean faster rates.

```{r}
rate.bin = findInterval(edge.rate, seq(0, max(edge.rate), length.out=33))
edge.color = colorRampPalette(
     c("#BEBEBE","#00008B","#FF0000","#FFA500","#FFD700"))(33)[rate.bin]
```

And we're ready to make another plot.

```{r}
plot(phy, edge.color=edge.color, lwd=0.5)
title("Model-averaged rates")
# Now highlight the nodes that underwent a state change
points(L[[1]][state.change, 1], L[[1]][state.change, 3], pch=21, bg=8, cex=0.8)
```

And what you should see is that all the regions of phylogeny that experienced 
lots of changes are inferred to have elevated rates of evolution compared to 
regions that underwent little to no change.

We can repeat the exact same process using individual rate-shift configurations
as well, not just the model-averaged rates. Let's look at the individual
configuration that has the highest score (highest AIC weight)

```{r}
best_index = which.max(fit$aic.weights)
```

Note that we have to be a little careful here due to different indexing
conventions between R and C. Rather than pass `best_index` directly, we need
to pass `best_index - 1`. This is why the names associated with the
`fit$aic.weights` vector are always one less than the vector position.

```{r}
edge.rate = fit$rate(best_index-1L)

# the particular rate-shift nodes can be recovered with
# attr(edge.rate, "shifts")

rate.bin = findInterval(edge.rate, seq(0, max(edge.rate), length.out=33))
edge.color = colorRampPalette(
     c("#BEBEBE","#00008B","#FF0000","#FFA500","#FFD700"))(33)[rate.bin]
plot(phy, edge.color=edge.color, lwd=0.5)
points(L[[1]][state.change, 1], L[[1]][state.change, 3], pch=21, bg=8, cex=0.8)
title("Best rate-shift configuration rates")
```

The overall pattern here is quite similar to the model-averaged pattern.

Finally, let's just plot the overall distribution of relative likelihood
scores.

```{r}
plot(fit$aic.weights, type='h', xlim=c(0, 100), las=1, ylab="Akaike weight",
    xlab="Model index")
```

Each model corresponds to a rate-shift configuration, and they are indexed by
the number of branches in the tree that belong to the background process. We
can see from this figure that there are two rate-shift configurations that
receive high support.

The weight assigned to `fit$aic.weights[1]` represents the relative likelihood
that the data were generated by a homogeneous rate process and it is a good
idea to always look at this value in relation to others before working with
model-averaged branch rates.

In this case, it is very unlikely that the data were generated by a constant
rate process:

```{r}
fit$aic.weights[best_index] / fit$aic.weights[1]
```

```{r echo=FALSE}
x = fit$aic.weights[best_index] / fit$aic.weights[1]
```

The best rate-shift configuration is roughly `r floor(log10(x))` orders of magnitude more likely.
